Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-05-10 manuscript.Rmd
Updates

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes]. This is shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for the interdependence between observations [@van2021bayes; @van2023bayes].} As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

## Log-normal Model 
A normal model may not be appropriate in case of more complex theories, hypotheses, or data structures. For example, the data in the example are response times. \Revision{Response times cannot be negative. The normal model, however, can yield negative values. When using the normal model for response times, we are assigning probability to negative values that are not possible. This seems very problematic.} In addition, RT distributions are right-skewed. A better representation of the data at hand is therefore given by a log-normal model [@schramm2019reaction]. This means that Equation \@ref(eq:datalevel) has to be adjusted:

@@

@@

Another important issue is that wide priors are often improper probability distributions [@hobert1996effect]. \Revision{A prior is proper if two conditions hold: 1. All values of the probability distribution are positive definite, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure \@ref(fig:properfigure). The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive definite values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

## Step 3: Prior Predictions 
A key advantage of model specification with priors is that we can obtain predictions on what the data would look like that are generated from the model \Revision{[@etz2018bayesian; @lee2018determining]}. These prior predictions allow us to make the implications of our specified priors more concrete. Researchers can look at relevant statistic that summarize the simulated data and decide if they correspond to reasonable observations for their specific research context. If the simulated data are in line with the researchers' expectations, the priors can be used for the analysis. If the simulated data do not match the expectations, the priors have to be adjusted (Step 4).

@@

### Model Fitting 
The model is fitted using the packages `rstan` and `brms`. \Revision{We used different settings for `rstan` and `brms` due to the difference between the models for these packages. Although we tried to make the models as similar as possible, the models are not identical. This is partly due to differences in prior specification (i.e., specifying prior on the variances versus standard deviations).} For `rstan`, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For `brms`, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

#### Number of Effective Samples 
Another popular convergence statistic is the number of effective samples (ESS). This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration *i* are similar to the parameter values at iteration *i-1*. The number of effective samples represents the estimated total number of *independent* draws from the posterior for every model parameter [@stan2018b, section 15.4]. If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. \Revision{A common rule for interpretation is that the number of effective samples should equal at least 100 per chain} [@vehtari2019rank]. \Revision{In the case of four chains, this would mean a value of 400 or higher. The higher the value the better. To get a better understanding of convergence, Vehtari and colleagues (2021) recommended evaluating different quantities of the posterior distribution instead, using so-called Bulk- and Tail-ESS. Bulk-ESS evaluates the center of the posterior distribution, while Tail-ESS evaluates the tails of the posterior distribution. For this tutorial, we focus on the basic calculation of the number of effective samples suggested by Gelman et al. (2013) and improved by Vehtari et al. (2021).}

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \Revision{The difference between the model diagnostics of `rstan` and `brms` can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). Although we tried to make the models as similar as possible, they are not identical.}

@@

Imagine this results in $\text{BF}_{\text{nl}} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 fulfills?}. As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

Same as the posterior distribution, Bayes factors can rarely be computed analytically due to the multidimensional integrals involved. The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect.

@@

\Revision{Note that models models 2-4 are nested in the most complex full model, and the null model is nested in models 2 and 3, but models 2 and 3 are not nested}. Additionally, the models differ by several parameters. For example, compared to the full model the Digit model restricts the side parameter for all participants to zero. Using bridge sampling, we can estimate the marginal likelihood for each model separately, and then compute Bayes factors as their ratios (Equation 14).

@@

We assess the stability of the obtained Bayes factor by repeating the estimating process ten times with `rstan` and `brms`, respectively. The results are shown in Figure \@ref(fig:bsbfestimatesbrmsrstan). The figure shows the evidence in favor of the null, digit and side model against the full model. All Bayes factors are larger than one. Therefore, it can be concluded that the data are more likely under less complex models. There is almost no variation in the computed Bayes factors per model, indicating that the Bayes factor estimate is stable. However, the Bayes factor estimates in `rstan` generally show stronger evidence than Bayes factor estimates based on `brms`. It is important to note that this difference is not due to the bridge sampling function, but a result of a difference in parametrization. \Revision{The models for `rstan` and `brms` are similar but not identical, partly due to the difference in prior specification (i.e., variances versus standard deviations).} However, for both packages, the conclusion is the same: the data are least likely to have occurred under the full model. In case the estimates are unstable, it is recommended to further increase the number of posterior samples.

@@

First, we kept the discussion of the software packages and code fairly brief. We provide further code examples for the digit classification task in the [Online Supplement](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment). For different models, it may be useful to study additional tutorials and books. For `rstan`, we recommend the `Stan` manual [@stan2018a; @stan2018b]. For `brms` there are several useful tutorials [@R-brms_a; @R-brms_b; @burkner2019ordinal]. There are also several resources for Bayesian estimation with other samplers [@lee2013bayesian; @rouder2013hierarchical]. Two other useful tutorials for Bayesian modeling are Schad, Betancourt, and
Vasishth [-@schad2021toward] and Schad, Nicenboim, Bürkner, Betancourt, and Vasishth [-@schad2021workflow]. \Revision{There are also other R software packages for Bayesian multilevel modeling besides `brms` and `rstan`, such as `BayesFactor`[@R-BayesFactor] and `rstanarm`[@rstanarm2020], that we did not illustrate in this tutorial. However, most of these other R software packages are a bit more limited, for instance, in model specification and prior settings.}

@@

Another technical issue may arise with the estimation of Bayes factors. Here, we highlighted two approaches, the Savage-Dickey density ratio [@wagenmakers2010bayesian] and bridge sampling [@gronau2017tutorial]. We already highlighted the drawbacks of the SD density ratio. \Revision{Bayes factor estimates can also be biased. However, bridge sampling has shown to be relatively unbiased [@sarafoglou2021evaluating].} Yet, bridge sampling may also fail in some cases. Failures of the bridge sampling algorithm are identified by unstable Bayes factors across repeated sampling runs. There are improvements to the algorithm that are specifically developed for hierarchical models [@gronau2019simple]. Yet, it remains crucial to assess the stability of Bayes factor estimates.

>>>>

Before diving into hierarchical modeling, let us consider two alternative approaches: Analyzing aggregated data and analyzing the data separately for each participant. Analyses of aggregated data such as the ANOVA from the scenario above examine a general effect that is assumed to be consistent across subjects. For instance, when investigating whether there is an effect of digit on RT, the mean RT for each participant is computed for each digit-condition, and a repeated-measures ANOVA is conducted on these aggregated scores. Analyses of aggregated data cannot investigate whether a trend is consistent across participants. Therefore, when this option is applied and individual variability is ignored, the general trend can be an inaccurate representation of the true general trend [Rouder, Morey, & Pratte, 2013; Haaf & Rouder, 2017; van Doorn, Aust, Haaf, Stefan, and Wagenmakers, -@van2021bayes]. This is shown in Panel A of Figure \@ref(fig:figure1) where some of the individual trends (i.e., slopes of the grey lines) differ from the general trend (i.e., slope of the blue line). If every participant takes part in each condition, the person variability affects all conditions, thus inducing a correlation across conditions. For example, participants in a digit classification task repeatedly respond to target digits inducing a correlation between the responses to the target digits, such as between target digit 3 and target digit 4. \Revision{Therefore, analyses on aggregated data do not appropriately account for the interdependence between observations} [@van2021bayes; @van2023bayes]. As a result, the type I error rate of a test of the effect increases: there is a higher probability of concluding that there is an effect if there is no effect (Rouder & Lu, 2005).

@@

## Log-normal Model 
A normal model may not be appropriate in case of more complex theories, hypotheses, or data structures. For example, the data in the example are response times. \Revision{Response times cannot be negative. The normal model, however, can yield negative values. When using the normal model for response times, we are assigning probability to negative values that are not possible. This could be problematic.} In addition, RT distributions are right-skewed. A better representation of the data at hand is therefore given by a log-normal model [@schramm2019reaction]. This means that Equation \@ref(eq:datalevel) has to be adjusted:

@@

\Revision{Up until now, we specified models without a correlation between the individual effects. This means, for instance, that we do not expect that the estimations for the individual effects $\delta_{7, i}$ and $\delta_{3, i}$ to be related. However, in cases where we would expect the estimation of individual effects to be related, we could estimate the correlation coefficient between individual effects. This would add complexity to the model. For the sake of the tutorial, we will focus on the models without individual effect correlation, but we illustrate the estimation of the model with correlation between the individual effects in} [Online Supplement C and D](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment).

@@

Another important issue is that wide priors are often \Revision{improper probability distributions} [@hobert1996effect]. \Revision{A prior is proper if two conditions hold: 1. All values of the probability distribution are positive definite, and 2. the probability distribution sums to one for discrete data or integrates to one for continuous data. An example of a proper and improper probability distribution are provided in Figure} \@ref(fig:properfigure). \Revision{The figure shows two hypothetical priors for parameter $\theta$. The first condition can be assessed by checking the y-values---both priors have only positive definite values as function output (0.1, 0.2, and 0.4). To assess the second condition, we need to sum up the function values. Panel A shows a proper distribution where the values sum to 1; panel B shows an improper distribution where the values sum to 1.2.}

@@

## Step 3: Prior Predictions 
A key advantage of model specification with priors is that we can obtain predictions on what the data would look like that are generated from the model [@etz2018bayesian; @lee2018determining]. These prior predictions allow us to make the implications of our specified priors more concrete. Researchers can look at relevant statistic that summarize the simulated data and decide if they correspond to reasonable observations for their specific research context. If the simulated data are in line with the researchers' expectations, the priors can be used for the analysis. If the simulated data do not match the expectations, the priors have to be adjusted (Step 4).

@@

### Model Fitting 
The model is fitted using the packages `rstan` and `brms`. \Revision{We used different settings for} `rstan` \Revision{and} `brms` \Revision{due to the difference between the models for these packages. Although we tried to make the models as similar as possible, the models are not identical. This is partly due to differences in prior specification (i.e., specifying prior on the variances versus standard deviations).} For `rstan`, four chains are run with 4000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 12000 samples from the joint posterior distribution. For `brms`, four chains are run with 6000 iterations per chain, of which 1000 serve as warm-up. This results in a total of 20000 draws from the joint posterior distribution.

@@

#### Number of Effective Samples 
Another popular convergence statistic is the number of effective samples (ESS). This statistic is concerned with the dependence of posterior samples. In MCMC algorithms, the samples are to some degree dependent on one another: the parameter values at iteration *i* are similar to the parameter values at iteration *i-1*. The number of effective samples represents the estimated total number of *independent* draws from the posterior for every model parameter [@stan2018b, section 15.4]. If the number of effective samples is low, this indicates a high dependency of iterations in the sampling procedure. Therefore, the number of effective samples can be seen as a measure of the amount of new information about the posterior distribution that is provided by the total number of samples drawn from the posterior. \Revision{A common rule for interpretation is that the number of effective samples should equal at least 100 per chain} [@vehtari2019rank]. \Revision{In the case of four chains, this would mean a value of 400 or higher. The higher the value the better. To get a better understanding of convergence, Vehtari and colleagues (2021) recommended evaluating different quantities of the posterior distribution instead, using so-called Bulk- and Tail-ESS. Bulk-ESS evaluates the center of the posterior distribution, while Tail-ESS evaluates the tails of the posterior distribution. For this tutorial, we focus on the basic calculation of the number of effective samples suggested by Gelman et al.} [-@gelman2013bayesian] \Revision{and improved by Vehtari et al.} [-@vehtari2019rank].

For the symbolic distance effect, the estimated number of effective samples is provided in panel D of Figure \@ref(fig:traceplots3p). For the package `rstan` the number is generally high, much higher than the actual number of samples depicted by the green dashed line. The `brms` package shows much more variation. The lowest effective sample sizes can be found for the intercept parameters. Table 1 also displays the number of effective samples for the general effects parameters in the `rstan` model. If the number of effective samples is too low, `brms` and `rstan` will return a warning. Here, this was not case, and all other diagnostics were good, so we continue with the interpretation of the estimates. \Revision{The difference between the model diagnostics of} `rstan` \Revision{and} `brms` \Revision{can be partly explained by the difference in model parametrization, such as the difference in the prior specification (i.e., specifying priors on variances versus priors on standard deviations). Although we tried to make the models as similar as possible, they are not identical.}

@@

Imagine this results in $\text{BF}_{\text{nl}} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 fulfills?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

Same as the posterior distribution, Bayes factors can rarely be computed analytically due to the multidimensional integrals involved. The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect. \Revision{The first method that we will discuss, Savage-Dickey density ratio, evaluates the general effects with the individual effects in the model, while the second method, bridge-sampling, evaluates the full model, the general and individual effects are considered together. We want to make you aware that there have been discussions on which method should be preferred without consensus} [@van2021bayes; @van2023bayes; @rouder2023principles; @singmann2023statistics]. \Revision{Therefore, we will illustrate the use of the two approaches and discuss their potential.}

@@

\Revision{Note that models 2-4 are nested in the most complex full model, and the null model is nested in models 2 and 3, but models 2 and 3 are not nested}. Additionally, the models differ by several parameters. For example, compared to the full model the Digit model restricts the side parameter for all participants to zero. Using bridge sampling, we can estimate the marginal likelihood for each model separately, and then compute Bayes factors as their ratios (Equation 14).

@@

We assess the stability of the obtained Bayes factor by repeating the estimating process ten times with `rstan` and `brms`, respectively. The results are shown in Figure \@ref(fig:bsbfestimatesbrmsrstan). The figure shows the evidence in favor of the null, digit and side model against the full model. All Bayes factors are larger than one. Therefore, it can be concluded that the data are more likely under less complex models. There is almost no variation in the computed Bayes factors per model, indicating that the Bayes factor estimate is stable. However, the Bayes factor estimates in `rstan` generally show stronger evidence than Bayes factor estimates based on `brms`. It is important to note that this difference is not due to the bridge sampling function, but a result of a difference in parametrization. \Revision{The models for} `rstan` \Revision{and} `brms` \Revision{are similar but not identical, partly due to the difference in prior specification (i.e., variances versus standard deviations).} However, for both packages, the conclusion is the same: the data are least likely to have occurred under the full model. In case the estimates are unstable, it is recommended to further increase the number of posterior samples.

@@

First, we kept the discussion of the software packages and code fairly brief. We provide further code examples for the digit classification task in the [Online Supplement](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment). For different models, it may be useful to study additional tutorials and books. For `rstan`, we recommend the `Stan` manual [@stan2018a; @stan2018b]. For `brms` there are several useful tutorials [@R-brms_a; @R-brms_b; @burkner2019ordinal]. There are also several resources for Bayesian estimation with other samplers [@lee2013bayesian; @rouder2013hierarchical]. Two other useful tutorials for Bayesian modeling are Schad, Betancourt, and
Vasishth [-@schad2021toward] and Schad, Nicenboim, Bürkner, Betancourt, and Vasishth [-@schad2021workflow]. \Revision{There are also other R software packages for Bayesian multilevel modeling besides} `brms` \Revision{and} `rstan`\Revision{, such as} `BayesFactor`[@R-BayesFactor] \Revision{and} `rstanarm`[@rstanarm2020]\Revision{, that we did not illustrate in this tutorial. However, most of these other R software packages are a bit more limited, for instance, in model specification and prior settings.}

@@

Another technical issue may arise with the estimation of Bayes factors. Here, we highlighted two approaches, the Savage-Dickey density ratio [@wagenmakers2010bayesian] and bridge sampling [@gronau2017tutorial]. We already highlighted the drawbacks of the SD density ratio. \Revision{Bayes factor estimates can also be biased. However, bridge sampling has shown to be relatively unbiased} [@sarafoglou2021evaluating]. Yet, bridge sampling may also fail in some cases. Failures of the bridge sampling algorithm are identified by unstable Bayes factors across repeated sampling runs. There are improvements to the algorithm that are specifically developed for hierarchical models [@gronau2019simple]. Yet, it remains crucial to assess the stability of Bayes factor estimates.
